---
title: "Differential abundance testing with _Milo_"
author:
  - Emma Dann
  - Mike Morgan
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: miloR
vignette: |
  %\VignetteIndexEntry{Differential abundance testing with Milo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  message=FALSE
)
```

```{r setup, message=FALSE, warning=FALSE}
library(miloR)
library(SingleCellExperiment)
library(scater)
library(dplyr)
library(patchwork)
```

# Load data

For this vignette we will use the mouse gastrulation single-cell data from [Pijuan-Sala et al. 2019](https://www.nature.com/articles/s41586-019-0933-9). The dataset can be downloaded as a `SingleCellExperiment` object from the [`MouseGastrulationData`](https://bioconductor.org/packages/3.12/data/experiment/html/MouseGastrulationData.html) package on Bioconductor. To make computations faster, here we will download just a subset of samples, 6 samples at stage E7 and 6 samples at stage E7.5. 

This dataset has already been pre-processed and contains a `pca.corrected` dimensionality reduction, which was built after batch correction using [`fastMNN`](https://bioconductor.org/packages/release/bioc/vignettes/batchelor/inst/doc/correction.html).

```{r}
library(MouseGastrulationData)        
embryo_data = EmbryoAtlasData(samples = c(2,  3,  4,  6, 10, 14, 15, 19, 20, 30, 31, 32))
embryo_data
```

# Visualize the data

We recompute PCA and UMAP embedding for this subset of cells to visualize the data.

```{r}
embryo_data <- embryo_data[,apply(reducedDim(embryo_data, "pca.corrected"), 1, function(x) !all(is.na(x)))]
embryo_data <- runUMAP(embryo_data, dimred = "pca.corrected", name = 'umap')

plotReducedDim(embryo_data, colour_by="stage", dimred = "umap")
```

We will test for significant differences in abundance of cells between these stages of development, and the associated gene signatures.

# Differential abundance testing

## Create a Milo object

For differential abundance analysis on graph neighbourhoods we first construct a `Milo` object. This extends the [`SingleCellExperiment`](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) class to store information about neighbourhoods on the KNN graph. 

```{r}
embryo_milo <- Milo(embryo_data)
embryo_milo
```


## Construct KNN graph

We need to add the KNN graph to the Milo object. This is stored in the `graph` slot, in [`igraph`](https://igraph.org/r/) format. The `miloR` package includes functionality to build and store the graph from the PCA dimensions stored in the `reducedDim` slot. In this case, we specify that we want to build the graph from the MNN corrected PCA dimensions.

```{r}
embryo_milo <- buildGraph(embryo_milo, k = 30, d = 30, reduced.dim = "pca.corrected")
```

Alternatively, one can add a precomputed KNN graph (for example constructed with Seurat or scanpy) to the `graph` slot using the adjacency matrix, through the helper function `buildFromAdjacency`.

<!-- Alternatively, if you already have a KNN graph (for example constructed with Seurat/scanpy) you can add it from the adjacency matrix. -->

<!-- ```{r} -->
<!-- # ## Build up a mock SNN graph made with Seurat -->
<!-- # pca_df <- reducedDim(traj_milo, "PCA") -->
<!-- # rownames(pca_df) <- traj_milo$cell_id -->
<!-- # snn_graph <- FindNeighbors(pca_df)[["snn"]] -->
<!-- #  -->
<!-- # graph(traj_milo) <-  graph(buildFromAdjacency(snn_graph, k=10)) -->
<!-- ``` -->

# 1. Defining representative neighbourhoods

We define the neighbourhood of a cell, the index, as the group of cells connected by an edge in the KNN graph to the index cell. For efficiency, we don't test for DA in the neighbourhood of every cell, but we sample as indices a subset of representative cells, using a KNN sampling algorithm used by [Gut et al. 2015](https://www.nature.com/articles/nmeth.3545). 

For sampling you need to define a few parameters:

- `prop`: the proportion of cells to randomly sample to start with (usually 0.1 - 0.2 is sufficient)
- `k`: the k to use for KNN refinement (we recommend using the same k used for KNN graph building)
- `d`: the number of reduced dimensions to use for KNN refinement (we recommend using the same d used for KNN graph building)
- `refined` indicated whether you want to use the sampling refinement algorith, or just pick cells at random. The default and recommended way to go is to use refinement. The only situation in which you might consider using random instead, is if you have batch corrected your data with a graph based correction algorithm, such as [BBKNN](https://github.com/Teichlab/bbknn), but the results of DA testing will be suboptimal.

```{r}
traj_milo <- makeNhoods(traj_milo, prop = 0.1, k = 10, d=30, refined = TRUE)
```

Once we have defined neighbourhoods, it's good to take a look at how big the neighbourhoods are (i.e. how many cells form each neighbourhood). This affects the power of DA testing. We can check this out using the `plotNhoodSizeHist` function. Empirically, we found it's best to have a distribution peaking between 50 and 100. Otherwise you might consider rerunning `makeNhoods` increasing `k` and/or `prop` (here the distribution looks ludicrous because it's a small dataset).

```{r}
plotNhoodSizeHist(traj_milo)
```

# Counting cells in neighbourhoods

Now we have to count how many cells from each sample are in each neighbourhood. We need to use the cell metadata and specify which column contains the sample information.

```{r}
traj_milo <- countCells(traj_milo, meta.data = data.frame(colData(traj_milo)), sample="Sample")
```

This adds to the `Milo` object a `n \times m` matrix, where n is the number of neighbourhoods and $m$ is the number of experimental samples. Values indicate the number of cells from each sample counted in a neighbourhood. This count matrix will be used for DA testing.

```{r}
head(nhoodCounts(traj_milo))
```

# Differential abundance testing

Now we are all set to test for differential abundance in neighbourhoods. We implement this hypothesis testing in a generalized linear model (GLM) framework, specifically using the Negative Binomial GLM implementation in [`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html).

We first need to think about our experimental design. The design matrix should match samples to a condition of interest. In this case the `Condition` is the covariate we are going to test for.

```{r}
traj_design <- data.frame(colData(traj_milo))[,c("Sample", "Condition")]
traj_design <- distinct(traj_design)

traj_design
```

Milo uses an adaptation of the Spatial FDR correction introduced by [cydar](https://bioconductor.org/packages/release/bioc/html/cydar.html), which accounts for the overlap between neighbourhoods. Specifically, each hypothesis test P-value is weighted by the reciprocal of the kth nearest neighbour distance. To use this statistic we first need to store the distances between nearest neighbors in the Milo object.

```{r}
traj_milo <- calcNhoodDistance(traj_milo, d=30)
```

Now we can do the test, explicitly defining our experimental design. 

```{r}
da_results <- testNhoods(traj_milo, design = ~ Condition, design.df = traj_design)
```

This calculates a Fold-change and corrected P-value for each neighbourhood, which indicates wheather there is significant differential abundance between conditions.

```{r}
da_results %>%
  arrange(- SpatialFDR) %>%
  head() 
```

# Visualize neighbourhoods displaying DA

To visualize DA results, we build an abstracted graph of neighbourhoods that we can superimpose on the single-cell embedding.

```{r, fig.width=10, fig.height=6}
traj_milo <- buildNhoodGraph(traj_milo)

plotUMAP(traj_milo) + plotNhoodGraphDA(traj_milo, da_results, alpha=0.05) +
  plot_layout(guides="collect")
```


<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>













